Text embedding models yield high-resolution insights into conceptual knowledge from short multiple-choice quizzes

### 2024-02-17 paper/main.tex
WIP updates to results section for GLMM analyses

We hypothesized that a topic model trained on transcripts of the two lectures
should also capture the conceptual knowledge probed by each quiz question. If
indeed the topic model could capture information about the deeper conceptual
content of the lectures (i.e., beyond surface-level details such as particular
word choices), then we should be able to recover a correspondence between each
lecture and questions \textit{about} each lecture. Importantly, such a
correspondence could not solely arise from superficial text matching between
lecture transcripts and questions, since the lectures and questions often used
different words (Supp.~Fig.~\jaccard) and phrasings. Simply comparing the
average topic weights from each lecture and question set (averaging across time
and questions, respectively) reveals a striking correspondence (Supp.
Fig.~\topicWeights). Specifically, the average topic weights from Lecture~1 are
strongly correlated with the average topic weights from Lecture~1 questions
($r(13) = 0.809,~p < 0.001$, 95\% confidence interval (CI)~$= [0.633,~0.962]$),
and the average topic weights from Lecture~2 are strongly correlated with the
average topic weights from Lecture~2 questions ($r(13) = 0.728,~p = 0.002$,
95\% CI~$= [0.456,~0.920]$). At the same time, the average topic weights from
the two lectures are \textit{negatively} correlated with their non-matching
question sets (Lecture~1 video vs.~Lecture~2 questions: $r(13) = -0.547,~p =
0.035$, 95\% CI~$= [-0.812, ~-0.231]$; Lecture~2 video vs.~Lecture~1 questions:
$r(13) = -0.612,~p = 0.015$, 95\% CI~$= [-0.874,~-0.281]$), indicating that the
topic model also exhibits some degree of specificity. The full set of pairwise
comparisons between average topic weights for the lectures and question sets is
reported in Supplementary Figure~\topicWeights.

@@

While an individual lecture may be organized around a single broad theme at a
coarse scale, at a finer scale, each moment of a lecture typically covers a
narrower range of content. Given the correspondence we found between the
variabilities in topic expression across moments of each lecture and questions
from its corresponding set (Fig.~\ref{fig:topics}), we wondered whether the
text embedding model might additionally capture these conceptual relationships
at a finer scale. For example, if a particular question asks about the content
from one small part of a lecture, we wondered whether the text embeddings could
be used to automatically identify the ``matching'' moment(s) in the lecture. To
explore this, we computed the correlation between each question's topic weights
and the topic weights for each second of its corresponding lecture, and found
that each question appeared to be temporally specific
(Fig.~\ref{fig:question-correlations}). In particular, most questions' topic
vectors were maximally correlated with a well-defined (and relatively narrow)
range of timepoints from their corresponding lectures, and the correlations
fell off sharply outside of that range (Supp.~Figs.~\forcesCorrs,~\bosCorrs).
We also qualitatively examined the best-matching intervals for each question by
comparing the question's text to the text of the most-correlated parts of the
lectures (Supp.~Tab.~\matchTab). Despite that the questions were excluded from
the text embedding model's training set, in general we found (through manual
inspection) a close correspondence between the conceptual content that each
question probed and the content covered by the best-matching moments of the
lectures. Two representative examples are shown at the bottom of
Figure~\ref{fig:question-correlations}.

@@

The ability to quantify how much each question is ``asking about'' the content
from each moment of the lectures could enable high-resolution insights into
participants' knowledge. Traditional approaches to estimating how much a
student ``knows'' about the content of a given lecture entail computing the
proportion of correctly answered questions. But if two students receive
identical scores on an exam, might our modeling framework help us to gain more
nuanced insights into the \textit{specific} content that each student has
mastered (or failed to master)? For example, a student who misses three
questions that were all about the same concept (e.g., concept $A$) will have
gotten the same \textit{proportion} of questions correct as another student who
missed three questions about three \textit{different} concepts (e.g., $A$, $B$,
and $C$). But if we wanted to help these two students fill in the ``gaps'' in
their understandings, we might do well to focus specifically on concept $A$ for
the first student, but to also add in materials pertaining to concepts $B$ and
$C$ for the second student. In other words, raw ``proportion-correct'' measures
may capture \textit{how much} a student knows, but not \textit{what} they know.
We wondered whether our modeling framework might enable us to (formally and
automatically) infer participants' knowledge at the scale of individual
concepts (e.g., as captured by a single moment of a lecture).

We developed a simple formula (Eqn.~\ref{eqn:prop}) for using a participant's
responses to a small set of multiple-choice questions to estimate how much the
participant ``knows'' about the concept reflected by any arbitrary coordinate
$x$ in text embedding space (e.g., the content reflected by any moment in a
lecture they had watched; see \nameref{subsec:traces}). Essentially, the
estimated knowledge at coordinate $x$ is given by the weighted average
proportion of quiz questions the participant answered correctly, where the
weights reflect how much each question is ``about'' the content at $x$. When we
apply this approach to estimate the participant's knowledge about the content
presented in each moment of each lecture, we can obtain a detailed time course
describing how much ``knowledge'' the participant has about the content
presented at any part of the lecture. As shown in
Figure~\ref{fig:knowledge-timeseries}A and C, we can apply this approach
separately for the questions from each quiz participants took throughout the
experiment. From just a few questions per quiz (see~\nameref{subsec:traces}),
we obtain a high-resolution snapshot (at the time each quiz was taken) of what
the participants knew about any moment's content, from either of the two
lectures they watched (comprising a total of 1,100 samples across the two
lectures).

@@

If we are able to accurately estimate a participant's knowledge about the
content tested by a given question, our estimates of their knowledge should
carry some predictive information about whether the participant is likely to
answer that question correctly or incorrectly. We developed a statistical
approach to test this claim. 
For each of the three quizzes, separately, we

%For each question, in turn, we used
%Equation~\ref{eqn:prop} to predict each participant's knowledge at the given
%question's embedding space coordinate, using all \textit{other} questions that
%participant answered on the same quiz. For each quiz, we grouped these
%predicted knowledge values into two distributions: one for the predicted knowledge at the
%coordinates of \textit{correctly} answered questions, and another for the
%predicted knowledge at the coordinates of \textit{incorrectly} answered
%questions (Fig.~\ref{fig:predictions}). We then used Mann-Whitney $\U$-tests to
%compare the means of these distributions of predicted knowledge.

@@

>>>>

We hypothesized that a topic model trained on transcripts of the two lectures
should also capture the conceptual knowledge probed by each quiz question. If
indeed the topic model could capture information about the deeper conceptual
content of the lectures (i.e., beyond surface-level details such as particular
word choices), then we should be able to recover a correspondence between each
lecture and questions \textit{about} each lecture. Importantly, such a
correspondence could not solely arise from superficial text matching between
lecture transcripts and questions, since the lectures and questions often used
different words (Supp.~Fig.~\jaccard) and phrasings. Simply comparing the
average topic weights from each lecture and question set (averaging across time
and questions, respectively) reveals a striking correspondence 
(Supp.~Fig.~\topicWeights). Specifically, the average topic weights from Lecture~1 are
strongly correlated with the average topic weights from Lecture~1 questions
($r(13) = 0.809,~p < 0.001$, 95\% confidence interval (CI)~$= [0.633,~0.962]$),
and the average topic weights from Lecture~2 are strongly correlated with the
average topic weights from Lecture~2 questions ($r(13) = 0.728,~p = 0.002$,
95\% CI~$= [0.456,~0.920]$). At the same time, the average topic weights from
the two lectures are \textit{negatively} correlated with the average topic weights from their non-matching
question sets (Lecture~1 video vs.~Lecture~2 questions: $r(13) = -0.547,~p =
0.035$, 95\% CI~$= [-0.812, ~-0.231]$; Lecture~2 video vs.~Lecture~1 questions:
$r(13) = -0.612,~p = 0.015$, 95\% CI~$= [-0.874,~-0.281]$), indicating that the
topic model also exhibits some degree of specificity. The full set of pairwise
comparisons between average topic weights for the lectures and question sets is
reported in Supplementary Figure~\topicWeights.

@@

While an individual lecture may be organized around a single broad theme at a
coarse scale, at a finer scale, each moment of a lecture typically covers a
narrower range of content. Given the correspondence we found between the
variabilities in topic expression across moments of each lecture and questions
from its corresponding set (Fig.~\ref{fig:topics}), we wondered whether the
text embedding model might additionally capture these conceptual relationships
at a finer scale. For example, if a particular question asks about the content
from one small part of a lecture, we wondered whether the text embeddings could
be used to automatically identify the ``matching'' moment(s) in the lecture. To
explore this, we computed the correlation between each question's topic weights
and the topic weights for each second of its corresponding lecture, and found
that each question appeared to be temporally specific
(Fig.~\ref{fig:question-correlations}). In particular, most questions' topic
vectors were maximally correlated with a well-defined (and relatively narrow)
range of timepoints from their corresponding lectures, and the correlations
fell off sharply outside of that range (Supp.~Figs.~\forcesCorrs,~\bosCorrs).
We also qualitatively examined the best-matching intervals for each question by
comparing the question's text to the transcribed text from the most-correlated parts of the
lectures (Supp.~Tab.~\matchTab). Despite that the questions were excluded from
the text embedding model's training set, in general we found (through manual
inspection) a close correspondence between the conceptual content that each
question probed and the content covered by the best-matching moments of the
lectures. Two representative examples are shown at the bottom of
Figure~\ref{fig:question-correlations}.

@@

The ability to quantify how much each question is ``asking about'' the content
from each moment of the lectures could enable high-resolution insights into
participants' knowledge. Traditional approaches to estimating how much a
student ``knows'' about the content of a given lecture entail administering some form of assessment (e.g., a quiz) and computing the
proportion of correctly answered questions. But if two students receive
identical scores on such an exam, might our modeling framework help us to gain more
nuanced insights into the \textit{specific} content that each student has
mastered (or failed to master)? For example, a student who misses three
questions that were all about the same concept (e.g., concept $A$) will have
gotten the same \textit{proportion} of questions correct as another student who
missed three questions about three \textit{different} concepts (e.g., $A$, $B$,
and $C$). But if we wanted to help these two students fill in the ``gaps'' in
their understandings, we might do well to focus specifically on concept $A$ for
the first student, but to also add in materials pertaining to concepts $B$ and
$C$ for the second student. In other words, raw ``proportion-correct'' measures
may capture \textit{how much} a student knows, but not \textit{what} they know.
We wondered whether our modeling framework might enable us to (formally and
automatically) infer participants' knowledge at the scale of individual
concepts (e.g., as captured by a single moment of a lecture).

We developed a simple formula (Eqn.~\ref{eqn:prop}) for using a participant's
responses to a small set of multiple-choice questions to estimate how much the
participant ``knows'' about the concept reflected by any arbitrary coordinate
$x$ in text embedding space (e.g., the content reflected by any moment in a
lecture they had watched; see \nameref{subsec:traces}). Essentially, the
estimated knowledge at coordinate $x$ is given by the weighted
proportion of quiz questions the participant answered correctly, where the
weights reflect how much each question is ``about'' the content at $x$. When we
apply this approach to estimate the participant's knowledge about the content
presented in each moment of each lecture, we can obtain a detailed time course
describing how much ``knowledge'' that participant has about the content
presented at any part of the lecture. As shown in
Figure~\ref{fig:knowledge-timeseries}A and C, we can apply this approach
separately for the questions from each quiz participants took throughout the
experiment. From just a few questions per quiz (see~\nameref{subsec:traces}),
we obtain a high-resolution snapshot (at the time each quiz was taken) of what
the participants knew about any moment's content, from either of the two
lectures they watched (comprising a total of 1,100 samples across the two
lectures).

@@

If we are able to accurately estimate a participant's knowledge about the
content tested by a given question, our estimates of their knowledge should
carry some predictive information about whether they are likely to answer that 
question correctly or incorrectly. We developed a statistical approach to test this claim. 
For each quiz question a participant answered, in turn, we used Equation~\ref{eqn:prop} to estimate their knowledge at the given question's embedding space coordinate, based on other questions that participant answered on the same quiz. 
We repeated this for all participants, and for each of the three quizzes. 
Then, separately for each quiz, we fit a generalized linear mixed model (GLMM) with a logistic link function to explain the likelihood of correctly answering a question as a function of estimated knowledge for its embedding coordinate, while accounting for random variation among participants and questions (see Sec.~\nameref{subsec:glmm}).
To assess the predictive value of the knowledge estimates, we performed likelihood-ratio tests comparing each GLMMs to an analogous (i.e., nested) ``null'' model that did not consider estimated knowledge.

@@

\subsubsection*{GLMM METHODS SECTION PLACEHOLDER}\label{subsec:glmm}
